---
title: "Census Income Project"
output: html_notebook
---


#Data Preprocessing:
##Replace all the missing values with NA.
```{r}
library(plotly)
```
##Let's take a look at the data structure
```{r}
str(censusData)
```

## convert column into character. Because the dataset was imported as Factor, we need to change the columns into characters so we can manipulate the data

```{r}
censusData$workclass<-as.character(censusData$workclass)
 censusData$occupation<-as.character(censusData$occupation)
 censusData$native.country<-as.character(censusData$native.country)
 censusData$education<-as.character(censusData$education)
 censusData$marital.status<-as.character(censusData$marital.status)
 censusData$relationship<-as.character(censusData$relationship)
 censusData$race<-as.character(censusData$race)
 censusData$sex<-as.character(censusData$sex)
 censusData$X<-as.character(censusData$X)
```

#Now, let's look at the new structure
```{r}
str(censusData)
```
##Let look at the missing value in the columns
   

```{r}
table(is.na(censusData))
censusData[censusData==" ?"]<- NA
```

##Remove all the rows that contain NA values.
```{r}
censusData <-na.omit(censusData)
```

##Remove all whitespaces from the columns.
```{r}
install.packages("stringr")
install.packages("dplyr")
library(stringr) 
library(dplyr)
censusData<-mutate_if(censusData, is.character, str_trim)
```

##In order to work with models, vizualize, we need to convert them into factors
```{r}
censusData$workclass<-as.factor(censusData$workclass)
censusData$occupation<-as.factor(censusData$occupation)
censusData$native.country<-as.factor(censusData$native.country)
censusData$education<-as.factor(censusData$education)
censusData$marital.status<-as.factor(censusData$marital.status)
censusData$relationship<-as.factor(censusData$relationship)
censusData$race<-as.factor(censusData$race)
censusData$sex<-as.factor(censusData$sex)
censusData$X<-as.factor(censusData$X)
```


```{r}
str(censusData)
```

#let's now plot
```{r}
plot(censusData$X)
```

##Data Manipulation: In this phase I will perform data manipulation to analyze the data set using various functions from the dplyr package
```{r}
summary(censusData)
```

##Extract the "education" column and store it in "census_ed"
```{r}
census_ed<-censusData$education
View(census_ed)
class(census_ed)
head(census_ed)
```

##Extract all the columns from "age" to "relationship" and store it in "census_seq".
```{r}
install.packages("dplyr")
library(dplyr)
census_seq<-select(censusData,age:relationship)
census_seq
```

##Extract the column number "5", "8", "11" and store it in "census_col"
```{r}
census_col<-censusData[,c(5,8,11)]
View(census_col)
head(census_col)
```

##Extract all the male employees who work in state-gov and store it in "male_gov".

```{r}
install.packages("dplyr")
library(dplyr)
male_gov<-censusData%>% filter(sex == "Male" & workclass=="State-gov")
View(male_gov)
```

##Extract all the 39 year olds who either have a bachelor's degree 
#    or who are native of United States and store the result in "census_us"
```{r}
table(censusData$native.country)
table(censusData$education)
census_us<-censusData%>%filter(age==39&(education=="Bachelors"|native.country=="United-States"))
View(census_us)
```

##Extract 200 random rows from the "census" data frame and store it in "census_200".

```{r}
census_200<-sample_n(censusData,200)
View(census_200)
```

##Get the count of different levels of the "workclass" column.

```{r}
install.packages("plyr")
library(plyr)
countWcls<-count(censusData$workclass)
countWcls
table(censusData$workclass)
```

##Calculate the mean of "capital.gain" column grouped according to "workclass".

```{r}
tapply(censusData$capital.gain,censusData$workclass,mean)
```

#Data Visualization:
```{r}
install.packages("ggplot2")
library(ggplot2)
```

##Build a bar-plot for the "relationship" column and fill the bars according to the "race"
# column.
```{r}
ggplot(censusData,aes(x=relationship,fill=race))+
  geom_bar()
```

##Set x-axis label to 'Categories of Relationships'
##Set y-axis label to 'Count of Categories'

```{r}
ggplot(censusData,aes(x=relationship,fill=race))+
  geom_bar()+
  labs(x="Categories of Relationships",y="Count of Categories")
```


##Fill the bars according to "sex"
```{r}
ggplot(censusData,aes(x=relationship,fill=sex))+
  geom_bar()+
  labs(x="Categories of Relationships",y="Count of Categories")
```

##Set the position of the bars to "dodge"

```{r}
ggplot(censusData,aes(x=relationship,fill=sex))+
  geom_bar(position = "dodge")+
  labs(x="Categories of Relationships",y="Count of Categories")
```

##Set the title of plot to be 'Distribution of Relationships by Sex"

```{r}
ggplot(censusData,aes(x=relationship,fill=sex))+
  geom_bar(position = "dodge")+
  labs(x="Categories of Relationships",y="Count of Categories",title = "Distribution of Relationships by Sex")
```

##Build a Histogram for the "age" column with number of bins equal to 50.
```{r}
ggplot(censusData,aes(x=age))+geom_histogram(bins = 50)
table(censusData$age)
```

##Fill the bars of the histogram according to yearly income column i.e., "X"
```{r}
ggplot(censusData,aes(x=age,fill=X))+geom_histogram(bins = 90)
```

##Set the title of the plot to "Distribution of Age".
```{r}
ggplot(censusData,aes(x=age,fill=X))+geom_histogram(bins = 90)+
  labs(title = "Distribution of Age")
```

##Set the legend title to "Yearly income".
```{r}
ggplot(censusData,aes(x=age,fill=X))+geom_histogram(bins = 90)+
  labs(title = "Distribution of Age",fill='Yearly income')
```

##Set the theme of the plot to black and white.

```{r}
ggplot(censusData,aes(x=age))+geom_histogram(bins =90)+
  labs(title = "Distribution of Age")+theme()
```

##Build a scatter-plot between "capital.gain" and "hours.per.week".
##     Map "capital.gain" on the x- axis and "hours.per.week" on the y-axis.

```{r}
ggplot(censusData,aes(x=capital.gain,y=hours.per.week))+geom_point()
```

##Set the transparency of the points to 40% and size as 2.

```{r}
ggplot(censusData,aes(x=capital.gain,y=hours.per.week))+
  geom_point(alpha=0.6,size=2)
```

##Set the color of the points according to the "X" (yearly income) column. 

```{r}
ggplot(censusData,aes(x=capital.gain,y=hours.per.week,fill=X))+geom_point()
```

##Set the x-axis label to "Capital Gain", y-axis label to "Hours per Week", title
# to "Capital Gain vs Hours per Week by Income", and legend label to "Yearly Income".

```{r}
ggplot(censusData,aes(x=capital.gain,y=hours.per.week,fill=X))+
  geom_point(alpha=0.6,size=2)+
labs(x="Capital Gain",y="Hours per Week",
     title = "Capital Gain vs Hours per Week by Income", fill="Yearly Income")
```


```{r}
install.packages("plotly")
library(plotly)
plot_ly(data=censusData, x = ~capital.gain, y = ~hours.per.week, color = ~X, type='scatter')
```

##Build a box-plot between "education" and "age" column.Map "education" on the x-axis and
## "age" on the y-axis.

 
```{r}
ggplot(censusData,aes(x=education,y=age))+geom_boxplot()
```

## Fill the box-plots according to the "sex" column.
 
```{r}
ggplot(censusData,aes(x=education,y=age,fill=sex))+geom_boxplot()
```

## Set the title to "Box-Plot of age by Education and Sex".
 
```{r}
ggplot(censusData,aes(x=education,y=age,fill=sex))+
   geom_boxplot()+labs(title = "Box-Plot of age by Education and Sex")
```

#Prediction guilding a Linear Regression Model:
## Build a simple linear regression model
##Divide the dataset into training and test sets in 70:30 ratio.
```{r}
 set.seed(98)
 install.packages("caTools")
 library("caTools")
 split_data<-sample.split(censusData$hours.per.week,SplitRatio = 0.70)
 View(split_data)
 censusTrain<-subset(censusData,split_data==T)
 censusTest<-subset(censusData,split_data==F)
 View(censusTrain)
 View(censusTest)
 nrow(censusTrain)
 nrow(censusTest)
View(split_data)
```

## Build a linear model on the train set where the dependent variable is
##"hours.per.week" and independent variable is "education.num". 
## dependent~independ
 
```{r}
View(censusData[c('hours.per.week','education.num')])
 
 LR_model<-lm(hours.per.week~education.num,data=censusTrain)
 summary(LR_model)
```
# Predicting the values on the test set and find the error in prediction. 
##Find the root-mean-square error (RMSE).


```{r}
 censusP<-predict(LR_model,newdata=censusTest)
 head(censusP)
 View(censusP)
 censusD<-cbind(Actual=censusTest$hours.per.week,Predicted=censusP)
 View(censusD)
 class(censusD)
 censusD<-as.data.frame(censusD)
 Error<-censusD$Actual-censusD$Predicted
 View(Error)
Data<-cbind(censusD,Error)
 View(Data)
```


```{r}
sqrt(mean((Data$Error)^2))
library(caret)
RMSE(censusP, censusTest$hours.per.week)
```

#Prediction building a 	Logistic Regression

## let's divide the dataset into training and test sets in 65:35 ratio.
 
```{r}
install.packages("caTools")
 library("caTools")
 split_data1<-sample.split(censusData$X,SplitRatio = 0.65)
 censusTrain1<-subset(censusData,split_data1==T)
 censusTest1<-subset(censusData,split_data1==F)
 nrow(censusTrain1)
 nrow(censusTest1)
```

### now let's build a logistic regression model where the dependent variable is "X"(yearly income) and independent variable is "occupation".

```{r}
log_mod<-glm(X~occupation,data=censusTrain1,family = "binomial")
summary(log_mod)
```

## Predicting the values on the test set.#### TO decide Accuracy we need to install ROCR. plot(acc)## Check for which valve accuracy get constant
```{r}
pred_val<-predict(log_mod,newdata =censusTest1,type = "response")#probability
head(pred_val)
range(pred_val)
install.packages("ROCR")
library(ROCR)
predict_log_roc<-prediction(pred_val,censusTest1$X)
predict_log_roc
acc<-performance(predict_log_roc,"acc")
plot(acc)
table(censusData$X)
```
## Plot accuracy vs cut-off and pick an ideal value for cut-off.

```{r}
lm.pred<-ifelse(pred_val>0.47,">50K","<=50K")  
lm.pred

```


# Build a confusion matrix and find the accuracy.

```{r}
tab<-table(lm.pred,censusTest1$X)
tab
```

#TP FP
#FN TN
#TP TN -correctly predicted
#FP FN - wrongly predicted


```{r}
(7188+660)/(7188+660+1968+741)
accuracy<-sum(diag(tab))/sum(tab)
accuracy
```

# Plot the ROC curve and find the auc(Area Under Curve). 

```{r}
install.packages("caTools")
 library("caTools")
roc<-performance(predict_log_roc,"tpr","fpr")
plot(roc)
performance(predict_log_roc, "auc")->auc
auc
auc<-auc@y.values[[1]]
auc
split_data1<- sample.split(censusData$X,SplitRatio = 0.80)
censusTrain2<-subset(censusData,split_data1==T)
censusTest2<-subset(censusData,split_data1==F)

log_mod2<-glm(X~age+workclass+education,data=censusTrain2,family = "binomial")
summary(log_mod2)
pred_val<-predict(log_mod2,newdata =censusTest2,type = "response")
head(pred_val)
```

##library(ROCR) ## TO decide Accuracy

```{r}
library(ROCR)
predict_log_roc<-prediction(pred_val,censusTest2$X)
predict_log_roc
acc<-performance(predict_log_roc,"acc")
plot(acc)
lm.pred<-ifelse(pred_val>0.45,">50K","<=50K")  
lm.pred

tab<-table(lm.pred,censusTest2$X)
tab
accuracy<-sum(diag(tab))/sum(tab)
accuracy

roc<-performance(predict_log_roc,"tpr","fpr")
plot(roc)
performance(predict_log_roc, "auc")->auc
auc
auc<-auc@y.values[[1]]
auc
```
#Prediction building a  Decision Tree Model:
## Divide the dataset into training and test sets in 70:30 ratio.
```{r}
set.seed(123)
install.packages("caTools")
 library("caTools")
split_data<-sample.split(censusData,SplitRatio = 0.70)
censusTrain<-subset(censusData,split_data==T)
censusTest<-subset(censusData,split_data==F)
nrow(censusTrain)
nrow(censusTest)
# Build a decision tree model where the dependent variable is "X"(Yearly Income) and the rest of the variables as independent variables
library(rpart)
library(rpart.plot) 

census_model<-rpart(formula = X~.,
                    data = censusTrain,
                    method = "class")
```

## Plot the decision tree 

```{r}
library(rpart)
library(rpart.plot)
rpart.plot(x= census_model, type= 5, extra = 0,tweak = 1.5)
```

## Predict the values on the test set
```{r}
class_prediction<-predict(census_model,
                          newdata = censusTest,
                          type = "class")
class_prediction
```

#TP FP
#FN TN
#TP TN -correctly predicted
#FP FN - wrongly predicted
# Build a confusion matrix and calculate the accuracy
```{r}
tab<-table(class_prediction,censusTest$X)
tab
sum(diag(tab))/sum(tab)
```
#Prediction building a random Forest:
## let's build a random forest model
## Divide the dataset into training and test sets in 80:20 ratio

```{r}
set.seed(123)
install.packages("caTools")
 library("caTools")
split_data<-sample.split(censusData$X,SplitRatio = 0.8)
censusTrain<-subset(censusData,split_data==T)
censusTest<-subset(censusData,split_data==F)
nrow(censusTrain)
nrow(censusTest)
```
## now let's build a random forest model where the dependent variable is "X"(Yearly Income) and the rest of the variables as independent variables and number of trees as 300.


```{r}
library(randomForest)

census_model<-randomForest(formula=X~.,
                           data=censusTrain,
                           ntree=300)

plot(census_model)
```


```{r}
text(census_model)
```
## Predict values on the test set

```{r}
cenus_prediction<-predict(census_model,
                          newdata = censusTest,
                          type = "class")
```

# Build a confusion matrix and calculate the accuracy
```{r}
tab<-table(cenus_prediction,censusTest$X)
tab
sum(diag(tab))/sum(tab)
```